Quantum Hoare Type Theory

### 2020-10-25 ms-paper/ms-paper.tex
Update all examples

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {emp} r : Bool {emp /\ Id(r, false)}
    = do q <= mkQbit false;
         return measQbit q
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {emp} r : Bool {emp /\ Id(r, false)}
-- P0: emp
    = do q <= mkQbit false;
-- P1: P0 \o (q |-> |0\>)
         return measQbit q
-- P2: P1 \o ((q |-> -) -o emp)
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {emp} r : Bool {emp}
    = do q <= mkQbit false;
         applyU (H q);
         return measQbit q
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {emp} r : Bool {emp}
-- P0: emp
    = do q <= mkQbit false;
-- P1: P0 \o (q |-> |0\>)
         applyU (H q);
-- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
         return measQbit q
-- P3: P2 \o ((q |-> -) -o emp)
\end{lstlisting}
\end{minipage}

\subsubsection{Testing Bell Pair}
\leavevmode
\todo{TODO: Should the type be \type{Bool} $\otimes$ \type{Bool}?}

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}  -- a,b \in {true, false}
         = do qa <= mkQbit false;
              applyU (H qa);
              qb <= mkQbit false;
              applyU (ifQ qa (X qb));
              return (measQbit qa, measQbit qb)
\end{lstlisting}
\end{minipage}

@@

\begin{lstlisting}[language=QHaskell]
testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
              -- P0: emp
         = do qa <= mkQbit false;
              -- P1: P0 \o (qa |-> |0\>)
              applyU (H qa);
              -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
              qb <= mkQbit false;
              -- P3: P2 \o (qb |-> |0\>)
              applyU (ifQ qa (X qb));
              -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
              return (measQbit qa, measQbit qb)
              -- P5: P4 \o ((qa |-> -) -o emp) \o ((qb |-> -) -o emp)
\end{lstlisting}

In the assertion P2, we show that only the portion of the quantum state referred to by the \texttt{applyU} command is affected. Similarly, in P4 we see that only the relevant portion of the heap containing the control and target qubits used in the CNOT operation are affected. In this case, we also needed to combine the two qubit state as they are now entangled. The last assertion, P5, merely encodes the strongest postcondition for measurement for each of the two qubits.

@@

\begin{turn}{90}
	\begin{minipage}{42pc}
	\begin{mathpar}
		\inferrule*[right=HIntro]
		{\inferrule*[right=consq]
			{\inferrule*[right=mkQbit]
				{\inferrule*[right=measQbit]
					{\inferrule*[right=return]
						{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap emp) \vdash \kw{return} r \syn r : \type{Bool} . R}
						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
						\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
					}
					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measQbit} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
					\\ \cdot \vdash \kw{false} \chk \type{Bool}
				}
				{
					\cdot; P_0 \vdash q \Leftarrow \kw{mkQbit false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
				}
				\\ r : \type{Bool}; i,m ; R \prove emp \multimap emp \wedge Id(r, \kw{false})
			}
			{
				\cdot; \lambda i . \lambda m . i = m \wedge (emp * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (emp \multimap emp \wedge Id(r, \kw{false}))
			}
		}
		{\cdot \vdash \kw{do} E \chk \{emp\} r : \type{Bool} \{emp \wedge \kw{Id}(r, \kw{false})\}}
	\end{mathpar}
	\end{minipage}
\end{turn}

@@

\subsection{Hadamard basis states}
\begin{lstlisting}[language=QHaskell]
qplus : {emp} r : Qbit {Id(r, |+\>)}
      = do q <= mkQbit false;
           applyU (H q);
           return q

qminus : {emp} r : Qbit {Id(r, |-\>)}
       = do q <= mkQbit true;
            applyU (H q);
            return q
\end{lstlisting}

\subsection{Creating entanglement}
\begin{lstlisting}[language=QHaskell]
share : \Pi a : Qbit.
        {a \in {|+\>, |-\>}}   -- a \in {b, c} is short for Id(a,b) \/ Id(a,c)
           b : Qbit
        {Id(a, b) /\ a \in {|0\>, |1\>}}
      = \a.do b <= mkQbit false;
              applyU (ifQ a (X b));
              return b
\end{lstlisting}

@@

\subsection{Testing modular Bell pair}
\begin{lstlisting}[language=QHaskell]
testBell : {emp} (a, b) : (Bool o* Bool) {emp /\ Id(a, b)}
         = do (qa, qb) <- bell;
              return (measQbit qa, measQbit qb)
\end{lstlisting}

\section{Teleportation}
\label{app:teleport}
\todo{TODO: Refine assertions}
\subsection{Alice's circuit}
\begin{lstlisting}[language=QHaskell]
alice : \Pi a : Qbit. \Pi e : Qbit
        {(Id(a, -) /\ entangled(e)}
           r : (Bool o* Bool)
        {emp}
      = \a.\e.do applyU (ifQ a (X e));
                 applyU (H a);
                 return (measQbit a, measQbit e)
\end{lstlisting}

\subsection{Bob's circuit} \todo{TODO: With \texttt{if ... then ... else} removed, this will not work}
\begin{lstlisting}[language=QHaskell]
bob : \Pi m1 m2 : Bool. \Pi e : Qbit
      {entangled(e)}
         r : Qbit
      {(Id(r, -)}
      = \m1.\m2.\e.do if m1 then applyU (Z e) else ();
                      if m2 then applyU (X e) else ();
                      return e
\end{lstlisting}

>>>>

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {emp} r : Bool {emp /\ Id(r, false)}
    = do q <= init false;
         return measure q
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {emp} r : Bool {emp /\ Id(r, false)}
-- P0: emp
    = do q <= init false;
-- P1: P0 \o (q |-> |0\>)
         return measure q
-- P2: P1 \o ((q |-> -) -o emp)
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {emp} r : Bool {emp}
    = do q <= init false;
         apply H to q;
         return measure q
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {emp} r : Bool {emp}
-- P0: emp
    = do q <= init false;
-- P1: P0 \o (q |-> |0\>)
         apply H to q;
-- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
         return measure q
-- P3: P2 \o ((q |-> -) -o emp)
\end{lstlisting}
\end{minipage}

\subsubsection{Testing Bell Pair}
\leavevmode
\todo{TODO: Should the type be \type{Bool} $\otimes$ \type{Bool}?}
\todo{TODO: syntax for applying multi-qubit gate}

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}  -- a,b \in {true, false}
         = do qa <= init false;
              apply H to qa;
              qb <= init false;
              apply CNOT to [qa, qb];
              return (measure qa, measure qb)
\end{lstlisting}
\end{minipage}

@@

\begin{lstlisting}[language=QHaskell]
testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
              -- P0: emp
         = do qa <= init false;
              -- P1: P0 \o (qa |-> |0\>)
              apply H to qa;
              -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
              qb <= init false;
              -- P3: P2 \o (qb |-> |0\>)
              apply CNOT to [qa, qb];
              -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
              return (measure qa, measure qb)
              -- P5: P4 \o ((qa |-> -) -o emp) \o ((qb |-> -) -o emp)
\end{lstlisting}

In the assertion P2, we show that only the portion of the quantum state referred to by the \texttt{apply} command is affected. Similarly, in P4 we see that only the relevant portion of the heap containing the control and target qubits used in the CNOT operation are affected. In this case, we also needed to combine the two qubit state as they are now entangled. The last assertion, P5, merely encodes the strongest postcondition for measurement for each of the two qubits.

@@

\begin{turn}{90}
	\begin{minipage}{42pc}
	\begin{mathpar}
		\inferrule*[right=HIntro]
		{\inferrule*[right=consq]
			{\inferrule*[right=init]
				{\inferrule*[right=measure]
					{\inferrule*[right=return]
						{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap emp) \vdash \kw{return} r \syn r : \type{Bool} . R}
						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
						\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
					}
					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measure} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
					\\ \cdot \vdash \kw{false} \chk \type{Bool}
				}
				{
					\cdot; P_0 \vdash q \Leftarrow \kw{init false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
				}
				\\ r : \type{Bool}; i,m ; R \prove emp \multimap emp \wedge Id(r, \kw{false})
			}
			{
				\cdot; \lambda i . \lambda m . i = m \wedge (emp * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (emp \multimap emp \wedge Id(r, \kw{false}))
			}
		}
		{\cdot \vdash \kw{do} E \chk \{emp\} r : \type{Bool} \{emp \wedge \kw{Id}(r, \kw{false})\}}
	\end{mathpar}
	\end{minipage}
\end{turn}

@@

\subsection{Hadamard basis states}
\begin{lstlisting}[language=QHaskell]
qplus : {emp} r : Qbit {Id(r, |+\>)}
      = do q <= init false;
           apply H to q;
           return q

qminus : {emp} r : Qbit {Id(r, |-\>)}
       = do q <= init true;
            apply H to q;
            return q
\end{lstlisting}

\subsection{Creating entanglement}
\begin{lstlisting}[language=QHaskell]
share : \Pi a : Qbit.
        {a \in {|+\>, |-\>}}   -- a \in {b, c} is short for Id(a,b) \/ Id(a,c)
           b : Qbit
        {Id(a, b) /\ a \in {|0\>, |1\>}}
      = \a.do b <= init false;
              apply CNOT to [a, b];
              return b
\end{lstlisting}

@@

\subsection{Testing modular Bell pair}
\begin{lstlisting}[language=QHaskell]
testBell : {emp} (a, b) : (Bool o* Bool) {emp /\ Id(a, b)}
         = do (qa, qb) <- bell;
              return (measure qa, measure qb)
\end{lstlisting}

\section{Teleportation}
\label{app:teleport}
\todo{TODO: Refine assertions}
\subsection{Alice's circuit}
\begin{lstlisting}[language=QHaskell]
alice : \Pi a : Qbit. \Pi e : Qbit
        {(Id(a, -) /\ entangled(e)}
           r : (Bool o* Bool)
        {emp}
      = \a.\e.do apply CNOT to [a, e];
                 apply H to a;
                 return (measure a, measure e)
\end{lstlisting}

\subsection{Bob's circuit}
\begin{lstlisting}[language=QHaskell]
bob : \Pi m1 m2 : Bool. \Pi e : Qbit
      {entangled(e)}
         r : Qbit
      {(Id(r, -)}
      = \m1.\m2.\e.do if m1 then (apply Z to e) else ();
                      if m2 then (apply X to e) else ();
                      return e
\end{lstlisting}
